Modelling Stochastic Time Delay for Regression Analysis

### 2021-04-18 documentation/article/article.tex
improve bib, fix table and add correct examples and remarks style
final comments on conclusion and future work

\theoremstyle{definition}
\newtheorem*{example}{Example}
\theoremstyle{definition}
\newtheorem*{remark}{Remark}

@@

\title{Modelling Stochastic Time Delay for Regression Analysis}
\author[Aaron]{Aaron Pickering}
\email{\href{mailto:aaron1rcl@gmail.com}{aaron1rcl@gmail.com}}
\author[Juan]{Juan Camilo Orduz}
\email{\href{mailto:juanitorduz@gmail.com}{juanitorduz@gmail.com}}
\urladdr{\href{https://juanitorduz.github.io/}{juanitorduz.github.io}}
%\address{}

\date{\today}
\maketitle

@@

@@

Therefore, we propose a regression model which can handle stochastic time delay structures. We treat the stochastic time delay components as an ‘error’ in the time axis. Next, we find the maximum likelihood estimate, for a given set of parameters, considering the time error ($t$-axis) and the regression error ($y$-axis) simultaneously.

@@

where $\theta_{f}$ and $\theta_{\mathcal{T}}$ represent the parameters of the model $f$ and time shifts $\mathcal{T}$ respectively. In other words, the $\mathcal{L}_2$ term represents likelihood of time shifts and the $\mathcal{L}_1$ term represents the likelihood of the prediction residuals. We maximise these terms simultaneously. For simplicity, we assume that the time shift distribution and error distributions are independent.

@@

For the model, we use the nomenclature Time Varying Stochastic (TVS) Regression and the full code to reproduce the example can be found via \href{https://github.com/aaron1rcl/tvs_regression/}{TVS Regression on GitHub}.

The following section demonstrates a simulated univariate example of TVS Regression. Figure \ref{fig:simulated_example} shows the simulated time series.

The input signal has 20 non-zero impulses which have been drawn from the standard normal distribution. The system is 'off' when $x(t) =0$. In other words, when $x(t) = 0$, $f(x(t)) = 0$.

The true shift distribution $\mathcal{T}$ is given by $\mathcal{T} \sim Poisson(\lambda = 2)$. There is one value of $\tau$ for each of the 20 impulses.

The green line represents the shifted series, corresponding to the time at which the effect occurs. The blue line is the output $y$ which includes a small amount of gaussian noise.

The values for the parameters were arbitrarily selected. The output y is therefore defined by the following equation:

\begin{align}
    y = 2(1^TX(t + \mathcal{T})) + 6.5
\end{align}

@@

\begin{center}
\begin{figure}
\includegraphics[scale=0.6]{images/convergence.png} 
\caption{Convergence}
\label{fig:convergence}
\end{figure}
\end{center}
\newpage
The true set of $\mathcal{T}$ is:

\begin{align}
\mathcal{T} = 
\left(
\begin{array}{cccccccccccccccccccc}
2 & 1 & 2 & 0 & 2 & 3 & 1 & 0 & 2 & 4 & 1 & 2 & 3 & 2 & 1 & 0  & 0 & 1 & 1 & 0 
\end{array}  
\right)
\end{align}
\newline

Compared to the estimated $\mathcal{T}$:

\begin{align}
\mathcal{T}_{est} = 
\left(
\begin{array}{cccccccccccccccccccc}   
2 & 1 & 2 & 0 & 2 & 3 & 1 & 1 & 2 & 4 & 1 & 2 & 3 & 2 & 1 & 0  & 0 & 1 & 1 & 0 
\end{array}  
\right)
\end{align}

\begin{center}
\begin{table}
\begin{tabular}{c|c|c|c|}
\hline
Parameter & TVS Regression & OLS Regression & True Value \\
\hline
$\beta$ & 2.09 & 0.50 & 2.00  \\
\hline
$Intercept$ & 6.53 & 6.62 & 6.50 \\
\hline
$\lambda_{\tau}$ & 1.54 & $N/A$ & 1.40 \\
\hline
$\sigma_{\epsilon}$ & 0.20 & 1.03 & 0.20 \\
\end{tabular}
\caption{ ... } 
\label{results_table}
\end{table}
\end{center}

As shown in Table \ref{results_table}, the estimated values for $\beta$ and $\sigma_{\epsilon}$ are significantly improved by taking into account the stochastic time delay noise. The estimated $\beta$ using OLS regression is $0.5$, compared to the true value $2$. Even a small amount of noise in the time axis causes attenuation, 
limiting the usefulness of OLS for these problems.
In comparison, the TVS Regression algorithm estimates $\beta$ to be $2.09$, much closer to the true value.

@@

\section{Future Work}

In the univariate case, the utility of the described method is limited. However, we propose three extensions as future work. The first is to extend the method to multiple regression. 
We believe that the extension can be built on the same fundamental ideas presented in this document. Next, the model could be extended to include distributed lag structures. A distributed lag structure is where past values of the impulse influence future values of the output \cite{almon}. Finally, in relating $x(t)$ to $f(x(t))$, the function $f$ could be extended to the modelling of non-linear forms.

Through these extensions we can begin to tackle a number of practical problems that are defined by stochastic relationships between the input and output e.g. the blood glucose problem described in the introduction.

In addition, by incorporating these elements we could begin to explore alternative forms of time series prediction. Essentially, existing prediction mechanisms predict the average across time shifts. However, in some situations the predictions could be considered nonsensical.

For example, let us consider a regression model to predict how much coffee someone drinks between 9am and 9:30am, at work. They set off every day in their car at exactly 8:30am and it takes about 30 minutes ($\pm \tau$) to get to work. The time shift $\tau$ could be caused by something as simple as variations in traffic. When they get to work, they always have a coffee. A typical regression might predict $0.5$ coffees between 8:30 am and 9am and $0.5$ coffees between 9am and 9:30am. Alternative prediction forms such as: 'When they get to work, they will have 1 coffee' might be more useful.

@@

We have proposed a form of regression analysis suited to the modelling of stochastic time delay problems. In addition, we have shown the feasibility of the approach and its performance on simulated data. Our approach allows for consistently improved estimation and prediction when the input is affected by noise in the time domain. To our knowledge, the method is novel for this class of problem.

%To-DO: Write References

\bibliography{references}
\bibliographystyle{ieeetr}

>>>>

\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\newtheorem{remark}{Remark}[section]

@@

\title{Modelling Stochastic Time Delay for Regression Analysis}
\author[Aaron Pickering]{Aaron Pickering}
\email{\href{mailto:aaron1rcl@gmail.com}{aaron1rcl@gmail.com}}
\author[Juan Camilo Orduz]{Juan Camilo Orduz}
\email{\href{mailto:juanitorduz@gmail.com}{juanitorduz@gmail.com}}
\urladdr{\href{https://juanitorduz.github.io/}{juanitorduz.github.io}}
%\address{}

\date{\today}

@@

\maketitle

@@

Therefore, we propose a regression model which can handle stochastic time delay structures. We treat the stochastic time delay components as an ‘error’ in the time axis. Next, we find the maximum likelihood estimate, for a given set of parameters, considering the time error ($t$-axis) and the regression error ($y$-axis) simultaneously.\\

An open source Python implementation is available on \href{https://github.com/aaron1rcl/tvs_regression/}{GitHub}.

@@

where $\theta_{f}$ and $\theta_{\mathcal{T}}$ represent the parameters of the model $f$ and time shifts $\mathcal{T}$ respectively. In other words, the $\mathcal{L}_2$ term represents likelihood of time shifts and the $\mathcal{L}_1$ term represents the likelihood of the prediction residuals. We maximise these terms simultaneously. For simplicity, we assume that the time shift distribution and error distributions are independent. \\

We will refer to this algorithm as {\em Time Varying Stochastic (TVS) Regression}.

@@

The following section demonstrates a simulated univariate example of TVS Regression. Figure \ref{fig:simulated_example} shows the simulated time series. The full code can be found via \href{https://github.com/aaron1rcl/tvs_regression/blob/master/notebooks/1_univariate_example.ipynb}{TVS Regression on GitHub}. 
The input signal has $20$ non-zero impulses which have been drawn from the standard normal distribution. The system is 'off' when $x(t) =0$. In other words, when $x(t) = 0$, $f(x(t)) = 0$. The true shift distribution $\mathcal{T}$ is given by $\mathcal{T} \sim Poisson(\lambda = 2)$. There is one value of $\tau$ for each of the 20 impulses. The green line in Figure \ref{fig:simulated_example} represents the shifted series, corresponding to the time at which the effect occurs. The blue line is the output $y$ which includes a small amount of gaussian noise. The values for the parameters were arbitrarily selected. The output $y$ is defined by the following equation:

\begin{align*}
    y = 2(1^TX(t + \mathcal{T})) + 6.5
\end{align*}

@@

\begin{center}
\begin{figure}
\includegraphics[scale=0.6]{images/convergence.png} 
\caption{Convergence}
\label{fig:convergence}
\end{figure}
\end{center}

The true values of $\mathcal{T}$ (true shifts) and the estimated values $\mathcal{T}_{est}$ are:

\begin{align*}
\mathcal{T} &= 
\left(
\begin{array}{cccccccccccccccccccc}
2 & 1 & 2 & 0 & 2 & 3 & 1 & 0 & 2 & 4 & 1 & 2 & 3 & 2 & 1 & 0  & 0 & 1 & 1 & 0 
\end{array}  
\right) \\
\mathcal{T}_{est} &= 
\left(
\begin{array}{cccccccccccccccccccc}   
2 & 1 & 2 & 0 & 2 & 3 & 1 & 1 & 2 & 4 & 1 & 2 & 3 & 2 & 1 & 0  & 0 & 1 & 1 & 0 
\end{array}  
\right)
\end{align*}

\begin{center}
\begin{table}
\begin{tabular}{|c|c|c|c|}
\hline
Parameter & TVS Regression & OLS Regression & True Value \\
\hline
$\beta$ & 2.09 & 0.50 & 2.00  \\
\hline
$Intercept$ & 6.53 & 6.62 & 6.50 \\
\hline
$\lambda_{\tau}$ & 1.54 & $N/A$ & 1.40 \\
\hline
$\sigma_{\epsilon}$ & 0.20 & 1.03 & 0.20 \\
\hline
\end{tabular}
\caption{ Parameter comparison between TVR and OLS regression against the true value.} 
\label{results_table}
\end{table}
\end{center}

As shown in Table \ref{results_table}, the estimated values for $\beta$ and $\sigma_{\varepsilon}$ are significantly improved by taking into account the stochastic time delay noise. The estimated $\beta$ using OLS regression is $0.5$, compared to the true value $2$. Even a small amount of noise in the time axis causes attenuation limiting the usefulness of OLS for these problems. In comparison, the TVS Regression algorithm estimates $\beta$ to be $2.09$, much closer to the true value.\\

@@

@@

We have proposed a form of regression analysis suited to the modelling of stochastic time delay problems. In addition, we have shown the feasibility of the approach and its performance on simulated data. Our approach allows for consistently improved estimation and prediction when the input is affected by noise in the time domain. To our knowledge, the method is novel for this class of problem.\\

We propose two extensions as future work. The first is to extend the method to multiple regression. We believe that the extension can be built on the same fundamental ideas presented in this document. Next, the model could be extended to include distributed lag structures. A distributed lag structure is where past values of the impulse influence future values of the output \cite{almon}. Through these extensions we can begin to tackle a number of practical problems that are defined by stochastic relationships between the input and output e.g. the blood glucose problem described in the introduction. In addition, we could also begin to apply the model to forecasting problems where the inputs are affected by stochastic time delay.

% References
\bibliographystyle{acm}
\bibliography{references}
